---
title: "coursework3"
author: "Bunmi"
date: "26/03/2021"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## INTRODUCTION

The data set consists of 2384 observations and 47 variables of police incidence which were obtained from January 2016 to January 2017.

This analysis would investigate police activities in Texas in the United State.It would explore the factors that leads to arrest,the crime rates and determine if there is an evidence of racial disparity in police activities .The following areas are going to be explored. 

* What is the monthly incident rate?
* Distribution of arrest by subject race. 
* Are traffic stop racially motivated?

 



```{r data ,echo=FALSE ,message =FALSE,include=FALSE}
#load the data and convert empty data to na 
mydata <- read.csv("../Assignment/37-00049_UOF-P_2016_prepped.csv",na.strings = c(""))
#remove the first row 
mydata <- mydata[-1,]
#convert empty Null to na 
mydata [mydata == "NULL"] <- NA
#view the data structure
str(mydata)
#change date format to date 
#tell r my date format first
mydata$INCIDENT_DATE <- as.Date(mydata$INCIDENT_DATE, format =  "%m/%d/%y")
#correct the date 
#get incident month in words by extracting from date
library(tidyverse)
mydata$INCIDENT_MONTH <- format(mydata$INCIDENT_DATE,"%m")
```

### Monthly Incident Rate


```{r  crime rate,message =FALSE,echo=FALSE}
#crime montly distribution
#library(tidyverse)
mydata$INCIDENT_MONTH <- format(mydata$INCIDENT_DATE,"%b")
month_G1 <- mydata %>%
  drop_na(INCIDENT_MONTH) %>%
  group_by(INCIDENT_MONTH)%>%
   summarise(Cout= n()) 
# #plot monthly incident to see if there is a pattern 

monthly_crime <- month_G1 %>% ggplot(aes(x=INCIDENT_MONTH,y=Cout))
z <- monthly_crime+ geom_line(stat="identity",color ="purple",group =1) + 
  geom_point(size = 3, color = "steelblue") +
labs(title= "Monthly Incident Count" ,y ="Number of Incidents" , x= "Months")
library(plotly)
ggplotly(z)

```



```{r  crime rate per race ,message =FALSE,echo=FALSE}
#crime montly distribution
#library(tidyverse)
mydata$INCIDENT_MONTH <- format(mydata$INCIDENT_DATE,"%m")
month_race <- mydata %>%
  group_by(INCIDENT_MONTH,SUBJECT_RACE)%>%
   summarise(Cout= n())%>% drop_na()
# #plot monthly incident to see if there is a pattern 

monthly_race_incid <- month_race  %>%ggplot(aes(x=INCIDENT_MONTH,y=Cout))+ geom_line(data= filter(month_race,SUBJECT_RACE=="Black"),stat="identity",group =1,size=1,aes(color="purple"))+geom_line(data= filter(month_race,SUBJECT_RACE=="White"),stat="identity",group =1,size=1,aes(color="coral"))+
geom_line(data= filter(month_race,SUBJECT_RACE=="Hispanic"),stat="identity",group =1,size=1,aes(color="blue"))
#incident for other races are relatively small 
#make r recognise color through guide and map legend
monthly_race_incid +scale_color_identity(name = "Subject Race",
                          breaks = c("purple", "coral", "blue"),
                          labels = c("Black", "White", "Hispanic"),
                          guide = "legend")+
labs(title= "Monthly Incident Count" ,y ="Number", x= "Months" )



```

```{r porportion of race arrest  during an incident,message =FALSE,warning=FALSE,fig.align='center',echo=FALSE}

library(scales)

library(plotly)


 sub_race_count <- table(factor(mydata$SUBJECT_WAS_ARRESTED),factor(mydata$SUBJECT_RACE))
 
 #find percentage of columns(subject race) and round up by 2 dp
sub_race <- round(100*prop.table(sub_race_count,2),2)
sub_race <-as.data.frame(sub_race)
sub_race <- sub_race  %>% rename(SUBJECT_WAS_ARRESTED =Var1,
                                 SUBJECT_RACE=Var2)
sub_race$label <- paste0(sub_race$label, "\n",
                         round(sub_race$Freq,2), "%")
# 
sub_race_count_plot <- sub_race %>% ggplot(aes(x=SUBJECT_RACE, y= Freq ,fill = SUBJECT_WAS_ARRESTED ))+ geom_bar(stat = "identity",position ="stack")+ 
 #geom_text(label = paste(Freq,"%"))+
 labs(title= "Percentage of Subject Race Incidents by arrest" ,y ="Percentage Incidents" , x= "Subject Race" , fill ="subject arrested")
  

ggplotly(sub_race_count_plot)
#plot the graph

```


```{r race likely to have an incident,message =FALSE,warning=FALSE,fig.align='center',echo=FALSE}


 
 #find percentage of columns(subject race) and round up by 2 dp
sub_race_arrest <- mydata %>% filter(SUBJECT_WAS_ARRESTED =="Yes") %>% group_by(SUBJECT_RACE)%>% summarise(number = n()) %>% 
  drop_na()%>%
  arrange(desc(number))%>%
  mutate(percent = 100*number/sum(number))
# 
sub_race_arrest_plot <- sub_race_arrest %>% ggplot(aes(reorder(x=SUBJECT_RACE,-percent), y= percent ))+ geom_bar(stat = "identity",fill= "yellow",color="blue")+ 
 #geom_text(label = paste(Freq,"%"))+
 labs(title= "Percentage of Arrested Subject by Race" ,y ="Percentage Incidents" , x= "Subject Race")
  

ggplotly(sub_race_arrest_plot)
#plot the graph

```



```{r incident reason count,message =FALSE,warning=FALSE,echo=FALSE  }
# cateogry data ,sum wouldnt work
#is there racial bias in Traffic stop
incid_rsn <- mydata %>% group_by(INCIDENT_REASON) %>% 
  count() %>%
  arrange(desc(n))



incid_rsn_plot <- incid_rsn %>% head(5) %>% 
  ggplot(aes(reorder(x=INCIDENT_REASON,n),y= n, fill=n))+
  geom_bar(stat = "identity", color = "black") +  
scale_fill_gradient(low="orange", high="blue")
incid_rsn_plot + labs(title = "Top Five Incident Reason", x= "Incident Reason",y="Count", fill= "Count")

```


```{r traffic stop pie plot,message =FALSE,warning=FALSE,echo=FALSE  }
# cateogry data ,sum wouldnt work
#is there racial bias in Traffic stop
library(scales) 
vehicle_stop <- mydata %>% 
  filter(INCIDENT_REASON == "Traffic Stop" ) %>%
  group_by(SUBJECT_RACE)%>%
  drop_na(SUBJECT_RACE)%>%
  summarise(number =n())%>%
  arrange(desc(SUBJECT_RACE)) %>% #calc prop
    mutate(prop = round(number * 100 / sum(number), 1),
           lab.ypos = cumsum(prop) - 0.5*prop)
#put label % line by line
vehicle_stop$label <- paste0(vehicle_stop$label, "\n",
                         round(vehicle_stop$prop,1), "%")

vehicle_stop_plot <- vehicle_stop %>% ggplot(aes(x=2,y=prop,fill= SUBJECT_RACE))+
  geom_bar(width =1, stat = "identity", color = "black") +  
coord_polar("y", start = 0,direction = 1)+
  geom_text(aes(y =lab.ypos ,label = label), color = "white") + xlim(0.5, 2.5) +
  theme_void()
vehicle_stop_plot +scale_fill_manual(values =c("plum2", "salmon1","gold2","maroon1"))
```




```{r ofcer histo by gender ,message =FALSE,warning=FALSE,echo=FALSE  }

#count number of officier years in the force
mydata$OFFICER_YEARS_ON_FORCE_cont <- as.numeric(mydata$OFFICER_YEARS_ON_FORCE)
library(plotly)
officer_yr_incid <- mydata %>% ggplot(aes(x=OFFICER_YEARS_ON_FORCE_cont)) +geom_histogram(fill= "bisque3",color="black",binwidth = 2)+
  scale_x_continuous(breaks = seq(0,36,2), limits = c(0,36))+
  theme_bw() +labs(title="Officer years in the force",y="Density",x="years")
ggplotly(officer_yr_incid) 

```



```{r ofcer yrs box plot ,message =FALSE,warning=FALSE,echo=FALSE  }
mydata$OFFICER_YEARS_ON_FORCE_cont <- as.numeric(mydata$OFFICER_YEARS_ON_FORCE)
officer_yr_incid_box <- mydata %>% ggplot(aes(x=OFFICER_RACE,y=OFFICER_YEARS_ON_FORCE_cont)) +
  geom_boxplot(notch=TRUE,fill= "cyan4",alpha =0.7)+
  theme_bw() 
ggplotly(officer_yr_incid_box) 

```



```{r subj weapon disp ,message =FALSE,warning=FALSE,echo=FALSE  }
#count number of first type of force used by officiers in the force
first_officer_force <- mydata %>% 
  count(TYPE_OF_FORCE_USED1)%>% 
  arrange(desc(n)) %>% 
  drop_na()

first_officer_force_plot <- first_officer_force %>% 
  head(10)%>% 
  ggplot(aes(reorder(x=TYPE_OF_FORCE_USED1,n), y= n))+ 
  geom_bar(stat="identity", color= "black" , fill="salmon1")+ 
  labs(title ="The top ten type of force used", x= "First Type of force used", y= "count")+ coord_flip() 

first_officer_force_plot + theme_bw()


```



```{r officer weapon disp ,message =FALSE,warning=FALSE,echo=FALSE  }
#TYPE OF FORCE USED ON SUBJECT RACE
force_on_sub_race <- mydata %>% filter(TYPE_OF_FORCE_USED1 =="Weapon display at Person" )%>% group_by(SUBJECT_RACE) %>%
summarise(number =n())%>%mutate(prop = round(number * 100 / sum(number), 1))%>%
drop_na()


#label
force_on_sub_race$percentage <- paste0(force_on_sub_race$percentage, "\n",
                         round(force_on_sub_race$prop,1), "%")

sub_type1_weap_drw <- force_on_sub_race %>%
  ggplot(aes(x=reorder(SUBJECT_RACE,number),y=percentage))+ geom_bar(stat = "identity", fill = "lightblue",  color = "black")
 sub_type1_weap_drw + geom_text(aes(label=percentage))+
labs(x = "Subject Race",y = "Percent", title  = "Weapon drawn by Police at Subject Race distribution")
 text <-mydata %>%filter(TYPE_OF_FORCE_USED1 =="Weapon display at Person" )%>%
   select(SUBJECT_RACE,SUBJECT_WAS_ARRESTED)
 #text
 #table(text$SUBJECT_WAS_ARRESTED,text$SUBJECT_RACE)
```



```{r 1st type of force ,message =FALSE,warning=FALSE,echo=FALSE  }
 #which  OFficer race is likely to draw weapon
 
 force_by_offic_race <- mydata %>% filter(TYPE_OF_FORCE_USED1 =="Weapon display at Person" )%>% group_by(OFFICER_RACE) %>%
summarise(number =n())%>%mutate(prop = round(number * 100 / sum(number), 1))%>%
drop_na()
#label a percentage column
force_by_offic_race$percentage <- paste0(force_by_offic_race$percentage, "\n",
                         round(force_by_offic_race$prop,1), "%")

force_by_offic_race_plot <- force_by_offic_race %>%
  ggplot(aes(x=reorder(OFFICER_RACE,number),y=percentage))+ geom_bar(stat = "identity",  color = "black",fill="coral")
 force_by_offic_race_plot + geom_text(aes(label=percentage))+
labs(x = "Officer Race",y = "Percent", title  = "Weapon drawn by Police distribution")
 
```

### Incident Injury 

The figures below shows the count of injury observed from both officer and subject race distribution.The first figure shows the officer injury by race .It can be seen that white officers had the highest count of injury at about 160 and the lowest was injury observed was from Asian officer which was notably small. 

The second figure shows the injury count of different race subject.The highest injury was observed from Black subjects at about 343 ,the second highest observed was from white subject and Asian subject had no injury.

```{r officer injury/sub injury ,message =FALSE,warning=FALSE,echo=FALSE  }
officer_hs <- mydata %>% count(OFFICER_INJURY)%>% 
  arrange(desc(n)) %>% 
  drop_na()

#race explanatory , injury response
sub_hs <- mydata %>%group_by(SUBJECT_RACE)%>%
  count(SUBJECT_INJURY)%>% 
  arrange(desc(n)) %>% 
  drop_na()
officier_injury <-ggplot(mydata,aes(x=OFFICER_RACE, fill = OFFICER_INJURY)) + geom_bar(position ="dodge")+
scale_y_continuous(breaks = seq(0,1500,250), limits = c(0,1500))
officier_injury+ labs(title = "Officer Race Injury Count", y= "Count", x="Officer Race")

subject_injury <- mydata %>% drop_na(SUBJECT_RACE)%>%
  arrange(desc(SUBJECT_RACE))%>%
  ggplot(aes(x=SUBJECT_RACE,fill= SUBJECT_INJURY,y=..count..))+geom_bar(position ="dodge")+
scale_y_continuous(breaks = seq(0,1000,250), limits = c(0,1000))
subject_injury+labs(title = "Subject Race Injury Count", y= "Count", x="Subject Race")

sub_inj <- ftable( mydata$SUBJECT_RACE~ mydata$SUBJECT_INJURY)
sub_inj_count <- round(prop.table(sub_inj,2),2)
sub_inj_count <- as.data.frame(sub_inj_count)


```


```{r crime ,message =FALSE,warning=FALSE,echo=FALSE,include=FALSE  }
sub_crime <- mydata %>% count(SUBJECT_OFFENSE)%>% 
  arrange(desc(n)) %>% 
  drop_na()%>% mutate(Total=cumsum(n))
pattern <- "^Drug Possession /|Drug Possession "
s <- c("Drug Possession  - Misdemeanor","Drug Possession  - Felony")
str_detect(s, "^Drug")

mydata$grp_crime <- mydata$SUBJECT_OFFENSE %>%
  str_replace("^Assa\\D+","Assualt") %>%
  str_replace("^Warr\\D+","Warrant")%>%
  str_replace("^Resisting Arrest\\D+|Resisting Arrest|Evading Arrest\\D+","Resisting Arrest") %>%
  
  str_replace("^Drug Possession\\D+|Drug Possession|Drug Possession[.-]|Man/Del/Dangerous Drug\\D+","Drug Possession ")%>%
  str_replace("^Traffic Violation\\D+|Traffic Violation","Traffic Violation") %>%
  str_replace("^Warrant/|Warrant","Warrant")%>%
  str_replace("^Burglary of a Vehicle\\D+|Burglary of a Vehicle|Burglary\\D+|Burglary","Burglary") %>%
  str_replace("^Public Intoxication\\D+|Public Intoxication ","Public Intoxication")%>%
  str_replace("^Other Misdemeanor Arrest\\D+|Other Misdemeanor Arrest","Misdemeanor Arrest") %>%
  str_replace("^Other Felony Arrest\\D+|Other Felony Arrest","Felony")%>%
  str_replace("^DWI\\D+|DWI|DWLI|DUI|DUI\\D+","Driving under influence") %>%
  str_replace("^Theft\\D+|Theft","Theft")%>%
  str_replace("^Fail to ID\\D+","Fail to ID")%>%
  str_replace("^Resisting Search\\D+|Resisting Search","Resisting Search") %>%
  str_replace("^Robbery\\D+","Robbery")%>%
  str_replace("^APOWW\\D+","APOWW")%>%
  str_replace("^Discharge Firearm\\D+","Discharge Firearm")%>%
  str_replace("^Unauthorize Use Vehicle\\D+","Unauthorize Use Vehicle") %>%
  str_replace("^Disorderly Conduct\\D+","Disorderly Conduct")%>%
  str_replace("^Crim Trespas\\D+","Tresspass")

#remove no arrest


```



```{r division incident,message =FALSE,warning=FALSE,echo=FALSE }
library(leaflet)
mydata$Latitude <- as.numeric(mydata$LOCATION_LATITUDE)
mydata$Longitude <- as.numeric(mydata$LOCATION_LONGITUDE)
#group incident by division
division_incid <-mydata %>% 
  group_by(DIVISION,SUBJECT_RACE)%>%
  summarise(number=n())%>%drop_na()%>%
  mutate(prop=100*round(number/sum(number),3))

division_incid_plot <-division_incid %>% ggplot(aes(x=DIVISION,y=prop,fill=SUBJECT_RACE))+
  geom_bar(stat="identity",color="grey") +coord_flip()+theme_bw()+
  scale_fill_manual(values=c("yellow","blue","cyan2","slategray1","pink","plum1"))
ggplotly(division_incid_plot)




```


```{r crime map ,message =FALSE,warning=FALSE,echo=FALSE  }


library(htmltools)
library(leaflet)
mydata$Latitude <- as.numeric(mydata$LOCATION_LATITUDE)
mydata$Longitude <- as.numeric(mydata$LOCATION_LONGITUDE)


mymap_crime <- leaflet(mydata) %>%
 addTiles() %>%
   addCircleMarkers(data = mydata[mydata$grp_crime=="Drug Possession ",], group = "Drugs",col="red",label = ~htmlEscape("Drugs"))%>%
  addCircleMarkers(data = mydata[mydata$grp_crime=="Theft",], group = "Theft",color="yellow",label = ~htmlEscape("Theft"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Assualt",], group = "Assualt",color="#fee090",label = ~htmlEscape("Assualt"))%>%
  addCircleMarkers(data = mydata[mydata$grp_crime=="Warrant",], group = "Warrant",color="purple",label = ~htmlEscape("Warrant"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Burglary",], group = "Burglary",color="orange",label = ~htmlEscape("Burglary"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Tresspass",], group = "Tresspass",color="peach",label = ~htmlEscape("Tresspass"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="APOWW",], group = "APOWW",color="light blue",label = ~htmlEscape("APOWW"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Robbery",], group = "Robbery",color="#4575b4",label = ~htmlEscape("Robbery"))%>%
addCircleMarkers(data = mydata[mydata$grp_crime=="Resisting Arrest",], group = "Resisting Arrest",color="#74add1",label = ~htmlEscape("Resisting Arrest"))


mymap_crime%>%  addLayersControl(
    baseGroups = c("OSM (default)", "Toner Lite"),
    #define layer group into overlay object
    overlayGroups = c("Theft","Drugs", "Assualt","Burglary","Tresspass","APOWW", "Warrant", "Resisting Arrest","Robbery"),
    #layer options to show one layer at a time
    options = layersControlOptions(collapsed = TRUE))


```

The map below shows the distribution of incident reason in the city of Dallas .The traffic stop distribution seems to be spread with a few concentration in certain parts of the city.It also shows a high distribution of arrest in the city .

```{r incident reason map ,message =FALSE,warning=FALSE,echo=FALSE  }
incid_rsn <- mydata %>% group_by(INCIDENT_REASON) %>% 
  count() %>%
  arrange(desc(n))

 

incident_map <-leaflet(mydata) %>%
  addTiles() %>% 
addCircleMarkers(lat=subset(mydata,INCIDENT_REASON=='Traffic Stop')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Traffic Stop')$Longitude ,color = "red",label = ~htmlEscape('Traffic Stop'),group='Traffic Stop') %>%
  addCircleMarkers(lat=subset(mydata,INCIDENT_REASON=='Arrest')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Arrest')$Longitude ,color = "blue",label = ~htmlEscape('Arrest'),group='Arrest')%>%
  addCircleMarkers(lat=subset(mydata, INCIDENT_REASON=='Call for Cover')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Call for Cover')$Longitude, color = "yellow",opacity = 0.2,label = ~htmlEscape('Call for Cover'),group= 'Call for Cover')%>%
  addCircleMarkers(lat=subset(mydata, INCIDENT_REASON=='Crime in Progress')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Crime in Progress')$Longitude, color = "purple",opacity = 0.2 ,label = ~htmlEscape('Crime in Progress'),group= 'Crime in Progress')%>%
  addCircleMarkers(lat=subset(mydata,INCIDENT_REASON=='Service Call')$Latitude, lng=subset(mydata, INCIDENT_REASON=='Service Call')$Longitude ,color = "pink",label = ~htmlEscape('Service Call'),group= 'Service Call')

incident_map %>%addLayersControl(
    baseGroups = c("OSM (default)", "Toner Lite"),
    #define layer group into overlay object
    overlayGroups = c('Traffic Stop','Arrest', 'Call for Cover',"Crime in Progress","Service Call"),
    #layer options to show one layer at a time
    options = layersControlOptions(collapsed = TRUE))

```

